Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-26 paper.tex

Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
a GISAID global ``all-time'' tree from Nextstrain.
We illustrate the backbone phylogeny by visualizing
a single tree in the middle of the viral genome
since other regions of the
genome show almost identical topologies.
It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
agreement with the Nextstrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendants form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).
Supplementary Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.

@@

\subsection{Recombinant Pango lineages}
\label{sec:pango_x_lineages}

In this section we focus on the detailed genealogy of samples that have been
previously identified as recombinants, i.e. given a Pango designation that
starts with an ``X''. We focus primarily on the Long ARG
since it contains many more recombinants. The status
of Pango X lineages in the Wide ARG is briefly
summarised in Section~\ref{sec-pango_x_wide_arg}.

The Long ARG contains 749 samples designated by Nextclade [CITATION] as Pango
recombinants, classed into 33 Pango lineages; this
reduces to 711 samples from the same 33 lineages
when singleton recombinants (i.e. recombination
nodes subtending only one sample) are removed.
In contrast, using the alternative GISAID designation process [CITATION], only
515 samples in the Long ARG are labelled as Pango recombinants,
which fall into 38 lineages (when singleton recombinants are excluded this
drops to 514 samples from 35 lineages). The variation in Nextclade versus
GISAID classifications highlights the frequent uncertainty that exists when
assigning Pango X lineages to samples.

Below we focus on the 749 Nextclade-designated recombinant samples, all
but two of which descend from a recombination node in the Long ARG.
These two samples are designated XP, and a likely
explanation for the absence of a corresponding recombination event in the Long ARG
is that the characteristic multibase deletion for XP
(\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
during preprocessing (See~\ref{sec:data_preprocessing} for details and potential
improvements). 38 of the remaining X designated samples are singleton recombinants,
descending
from recombination nodes that are ancestral only to that sample (10 are XZ; 6
are XE;  3 each from XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1
sample from each of XB, XM, XJ, XAF, XAH and XAJ). Such samples are likely to
be enriched for sequencing errors and lineage designation artefacts, and so we
exclude them from further analysis in this section.

The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, caused by strain
Denmark/DCGC-281594/2021, is ancestral to more than 127K samples,
and is highly likely to be a spuriously inferred recombination event.
Understanding the origin of this
and other similar high-impact, but likely spurious recombination
events is an important aspect of future work.
For simplicity, we exclude these samples from further analyses in
this section.

The remaining remaining 630 Pango recombinant samples trace back to 50 different
most recent recombination nodes, corresponding to 31
Nextclade-designated Pango lineages. These are summarised
in Supplementary Table~\ref{tab:pango-recombinants}, and fall
into three classes: single origin, multiple origin,
and multiple nested origins, which we discuss in the following sections.

\subsubsection{Single origin}
In the absence of genealogical information, a reasonable initial assumption is
that all sequences
assigned to a given Pango X lineage are descendants of a single recombinant
sequence, arising as a result of a mixed infection followed by onward
transmission. We would expect our ARG to reveal evolutionary histories of this
nature, where all the samples assigned to a given recombinant lineage
trace back to one recombination node, representing a single originating
recombination event. In the Long ARG, 16 of the 31 Pango recombinants lineages identified by Nextclade
fall into this category (Supplementary Table~\ref{tab:pango-recombinants})

@@

Six of the 16 Nextclade designated lineages (XA, XAC, XAE, XF, XK, and XS) are of the
basic (XA) type with no other Pango designations among their descendants.
The remaining ten are of the XAG type with multiple Pango
designated lineages as additional descendants of the originating recombination event.
In some cases,
these may be, however, be a result of erroneous Pango designations.
Supplementary Table~\ref{tab:pango-recombinants} also shows the official
Pango designated parent lineages and \texttt{sc2ts} inferred parent lineages,
which extensively agree (although \texttt{sc2ts} provides more a precise
parent designation).

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. Using Nextclade-designations, 15 Pango X lineages are
inferred to have multiple
recombinant origins, such that their samples trace to $>$1 most recent
recombination node in the ARG. Of these, 11 are cases where the
recombinants are independent rather than nested (i.e. there is no overlap in the list
of descendant samples for each recombination node). Most have a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend, and which agrees with the offical Pango designated parent lineages
(see Supplementary Table~\ref{tab:pango-recombinants}, but note that
in XJ and XU there are too few Pango X samples to decide on a ``main'' clade).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple multiple-origin example, consisting of
the 4 samples labelled XD by Nextclade in the Long ARG. The left hand subgraph
(the origin of three XD samples, all sampled in France) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
the breakpoints in each case are slightly different: the start of the Omicron insertion
in the French samples has an estimated maximum position of 21641 basepairs
(and minimum of 21619, not shown) with the insertion end occurring at a maximum position of 25584
(and a minimum of 25470). In contrast, the Omicron insertion in the Turkish sample
is inferred to have occurred between positions 22578 (min 21619) and 24130 (min 23605).
The breakpoint difference, the different geographical locations, the time between the
samples, and the fact that the two samples differ at 23 nucleotide positions,
suggests that these may indeed represent true independent Deltacron recombinants.
The canonical XD definition is based entirely on samples from northern Europe, particularly
France (\url{https://github.com/cov-lineages/pango-designation/issues/444}) so it seems
plausible that the earlier Turkish sample has been mislabelled XD by Nextclade.
Indeed, GISAID does not label any of these samples XD (see Supplementary
Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).
Investigation of other multiple-origin examples reveals somewhat similar patterns,
suggesting that most of the simple multiple origin examples are due to incorrect
Pango labelling.

\subsubsection{Nested recombinant origins}
As well as cases where Pango X lineage origins are attributed to independent
recombination events, four Pango X lineages in the Long ARG have Nextclade-designated
samples whose ancestry involves multiple non-independent (``nested'') recombination events
(marked by~\textdagger ~in Supplementary Table~\ref{tab:pango-recombinants}; the most
complex appears to be XAB). Figure~\ref{fig:complex_origins_graph} plots the earliest example,
XB,  which is present in both the Wide and Long ARGs. The subgraph shows a recombination
between a B.1 sample and B.1.627 sample that leads not only to all the XB-labelled samples
but also to a ``hairball'' of further recombination nodes whose descendants are often not
identified as recombinants by Nextclade (plotted on the left, in blue). A similar
pattern is seen when examining XB in the Wide ARG (see discussion below).

@@

Note that in the Long ARG, the nested recombination events account for only one
XB sample (pink upper left, with 7 mutations above it); moreover, this sample is
not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph})
indicating some uncertainty in lineage assignment in this part of the ARG.
Also note that the number of mutations on the lineages immediately above and
below the recombination node (totalling 18+2) is rather large, suggesting
that the sampled recombinant which induced the recombination node in the Long ARG
is only distantly related to the true originating recombinant. This could account
for complex and potentially artefactual relationships around these nodes, and
is likely to be due to undersampling of the XB outbreak. Investigating examples
of nested recombinant origins, and identifying which (if any) of the nested
recombination events may be artefactual, is an important area of future research.

>>>>

Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
a GISAID global ``all-time'' tree from Nextstrain.
We illustrate the backbone phylogeny by visualizing
a single tree in the middle of the viral genome
since other regions of the
genome show almost identical topologies.
It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
agreement with the Nextstrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendants form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).
Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.

@@

\subsection{Recombinant Pango lineages}
\label{sec:pango_x_lineages}
In this section we focus on the detailed ancestry of samples that have been
previously identified as recombinants, i.e., designated as belonging
to a Pango lineage with a name starting with an ``X''.
We focus primarily on the Long ARG
since it contains many more recombinants (the status
of Pango X lineages in the Wide ARG is briefly
summarised in Section~\ref{sec-pango_x_wide_arg}).
Designation of samples to Pango lineages is not a straightforward task,
and there can be significant variation between methods.
Here, we consider two different assignments of Pango lineages to samples in
the Long ARG, those provided by Nextclade and by GISAID.
Using the Nextclade assignments, the Long ARG contains 749 samples from
33 Pango X lineages (711 samples from 33 lineages when we
remove singleton recombinants).
In contrast, using the GISAID designations we
have 515 samples from 38 X lineages
[FIXME this doesn't make sense! How did we drop 1 sample and 3 lineages?]
(514 samples from 35 lineages when we filter singletons).
This variation in classifications highlights the uncertainty that exists when
assigning Pango X lineages to samples, and is important to keep in mind
when interpreting the results here.

We focus here on the 749 Nextclade-designated recombinant samples, all
but two of which descend from a recombination node in the Long ARG.
These two samples are designated XP, and a likely
explanation for the absence of a corresponding recombination event in the Long ARG
is that the characteristic multibase deletion for XP
(\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
during preprocessing (See Section~\ref{sec:data_preprocessing} for details and potential
improvements). Of the remaining X designated samples, 38 are singleton recombinants,
descending
from recombination nodes that are ancestral only to that sample (10 are XZ; 6
are XE;  3 each from XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1
sample from each of XB, XM, XJ, XAF, XAH and XAJ). Such samples are likely to
be enriched for sequencing errors and lineage designation artefacts, and so we
exclude them from further analysis in this section.

The most recent recombination node for 79 samples (XN: 53, XZ: 16, XAJ: 6, XE:
1, XAD: 1, XAH: 1, XAK: 1) is the same. This node, introduced as a
result of adding strain
Denmark/DCGC-281594/2021, is ancestral to more than 127K samples,
and is highly likely to be a spuriously inferred recombination event.
Understanding the origin of this
and other similar high-impact, but likely spurious recombination
events is an important aspect of future work.
For simplicity, we exclude these samples from further analyses in
this section.

The remaining 630 samples (31 Pango lineages) trace back to 50 different
most recent recombination nodes, summarised
in Table~\ref{tab:pango-recombinants}.
These fall into three classes: single origin, multiple origin,
and multiple nested origins, which we discuss in the following sections.

\subsubsection{Single origin}
In the absence of genealogical information, a reasonable initial assumption is
that all sequences
assigned to a given Pango X lineage are descendants of a single recombinant
sequence, arising as a result of a mixed infection followed by onward
transmission. We would expect our ARG to reveal evolutionary histories of this
nature, where all the samples assigned to a given recombinant lineage
trace back to one recombination node, representing a single originating
recombination event. In the Long ARG, 16 of the 31 Pango recombinants lineages identified by Nextclade
fall into this category (Table~\ref{tab:pango-recombinants})

@@

Six of the 16 Nextclade designated lineages (XA, XAC, XAE, XF, XK, and XS) are of the
basic (XA) type with no other Pango designations among their descendants.
The remaining ten are of the XAG type with multiple Pango
designated lineages as additional descendants of the originating recombination event.
In some cases,
these may be, however, be a result of erroneous Pango designations.
Table~\ref{tab:pango-recombinants} also shows the official
Pango designated parent lineages and \texttt{sc2ts} inferred parent lineages,
which extensively agree (although \texttt{sc2ts} provides more a precise
parent designation).

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. Using Nextclade-designations, 15 Pango X lineages are
inferred to have multiple
recombinant origins, such that their samples trace to $>$1 most recent
recombination node in the ARG. Of these, 11 are cases where the
recombinants are independent rather than nested (i.e. there is no overlap in the list
of descendant samples for each recombination node). Most have a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend, and which agrees with the offical Pango designated parent lineages
(see Table~\ref{tab:pango-recombinants}, but note that
in XJ and XU there are too few Pango X samples to decide on a ``main'' clade).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple multiple-origin example, consisting of
the 4 samples labelled XD by Nextclade in the Long ARG. The left hand subgraph
(the origin of three XD samples, all sampled in France) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
the breakpoints in each case are slightly different: the start of the Omicron insertion
in the French samples has an estimated maximum position of 21641 basepairs
(and minimum of 21619, not shown) with the insertion end occurring at a maximum position of 25584
(and a minimum of 25470). In contrast, the Omicron insertion in the Turkish sample
is inferred to have occurred between positions 22578 (min 21619) and 24130 (min 23605).
The breakpoint difference, the different geographical locations, the time between the
samples, and the fact that the two samples differ at 23 nucleotide positions,
suggests that these may indeed represent true independent Deltacron recombinants.
The canonical XD definition is based entirely on samples from northern Europe, particularly
France (\url{https://github.com/cov-lineages/pango-designation/issues/444}) so it seems
plausible that the earlier Turkish sample has been mislabelled XD by Nextclade.
Indeed, GISAID does not label any of these samples XD (see
Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).
Investigation of other multiple-origin examples reveals somewhat similar patterns,
suggesting that most of the simple multiple origin examples are due to incorrect
Pango labelling.

\subsubsection{Nested recombinant origins}
As well as cases where Pango X lineage origins are attributed to independent
recombination events, four Pango X lineages in the Long ARG have Nextclade-designated
samples whose ancestry involves multiple non-independent (``nested'') recombination events
(marked by~\textdagger ~in Table~\ref{tab:pango-recombinants}; the most
complex appears to be XAB). Figure~\ref{fig:complex_origins_graph} plots the earliest example,
XB,  which is present in both the Wide and Long ARGs. The subgraph shows a recombination
between a B.1 sample and B.1.627 sample that leads not only to all the XB-labelled samples
but also to a ``hairball'' of further recombination nodes whose descendants are often not
identified as recombinants by Nextclade (plotted on the left, in blue). A similar
pattern is seen when examining XB in the Wide ARG (see discussion below).

@@

Note that in the Long ARG, the nested recombination events account for only one
XB sample (pink upper left, with 7 mutations above it); moreover, this sample is
not identified as XB by GISAID (Figure~\ref{fig:pango_XB_gisaid_graph})
indicating some uncertainty in lineage assignment in this part of the ARG.
Also note that the number of mutations on the lineages immediately above and
below the recombination node (totalling 18+2) is rather large, suggesting
that the sampled recombinant which induced the recombination node in the Long ARG
is only distantly related to the true originating recombinant. This could account
for complex and potentially artefactual relationships around these nodes, and
is likely to be due to undersampling of the XB outbreak. Investigating examples
of nested recombinant origins, and identifying which (if any) of the nested
recombination events may be artefactual, is an important area of future research.
